Metropolitan Statistical Area Evaluation

### 2022-11-15 src/demo_data_wrangling.Rmd
cleaning up data (removed pay roll and tax info)

### Population Growth
All states were considered with the exception of CA, HI, and AK. Initial analysis of Metropolitan Statistical Areas (MSAs) relied on Census data of annual population estimates, in which the population had to be greater than 300,000 individuals for the purpose of attracting investors. Population change from 2020 to 2021 was calculated by $$\frac{2021 estimate- 2020 estimate}{2021 estimate} * 100$$
MSAs with a population growth > 0.5%, were selected to gather further data on. The 0.5% growth was determined.

@@

```{r population growth data wrangling}
# read in the data MSA population data from Census
msa_pop_2021_data <- read_csv(here("data","raw", "msa_2021_alldata.csv")) %>% 
  clean_names() %>% # data wrangling to make column headers in lower_snake_case
  filter(lsad == "Metropolitan Statistical Area") %>% # select MSA cities
  select("cbsa", "name", "popestimate2020", "popestimate2021", "residual2020", "residual2021") %>% # select data

  separate(col = "name",
           into = c("MSA", "state"), # split city and state data
           sep = ", ") %>% 
  mutate(pop_percent_change = (popestimate2021-popestimate2020)/popestimate2021 * 100)  # add population % change column

# create object with viable MSAs 
potential_MSAs <- msa_pop_2021_data %>% 
  filter(pop_percent_change > 0.5 ) %>% 
  mutate(pop_percent_change20_21 = round(pop_percent_change, digits = 3)) %>% 
  filter (!state %in% c("CA", "HI", "AK")) %>% 
  #separate(col = "state", into = c("state1", "state2", "state3"),    sep = "-") %>% 
  filter(!cbsa %in% c("10580", "10900", "12420", "12940", "14260", "15980", "18580", "19100", "19660", "19740", "23540", "25420", "26420", "27260", "29820", "34940", "36260", "36740", "38060", "39100", "39300", "39340", "41620", "45220", "45300", "46060", "46140", "39900")) %>%  # remove MSAs as instructed by Owen 
  mutate(normalized_pop2021 = (popestimate2021-min(popestimate2021))/ (max(popestimate2021)-min(popestimate2021))) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021)

@@

```{r fig.height = 7, fig.align='center', eval = FALSE}
ggplot(data = positive_MSA_pop_change, aes(x = pop_percent_change20_21, y = reorder(MSA, pop_percent_change20_21))) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_minimal(base_size = 12.5)

@@

Dataset: Median Household Income (USD) INCOME IN THE PAST 12 MONTHS (IN 2020 INFLATION-ADJUSTED DOLLARS)

@@

```{r household income data wrangling}
MSA_income_data <- read_csv(here("data","raw", "MSA_income", "2021_data", "2021_income_data.csv")) %>% 
  row_to_names(1) %>% 
  clean_names() %>% 
  mutate(median_income = as.numeric(estimate_households_median_income_dollars)) %>% 
  mutate(income_moe_USD = as.numeric(margin_of_error_households_median_income_dollars)) %>% 
  separate(col = "geography",
           into = c("id", 'cbsa'),
           sep = 9) %>% 
  mutate(cbsa = as.numeric(cbsa)) # change column class from chr to dbl

@@

avg_emp_yoy <- MSA_employment %>% 
  group_by(MSA, year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_YoY = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100) %>% 
  mutate(avg_emp_YoY = round(avg_emp_YoY, digits = 3)) %>% 
  drop_na()

@@

```{r}
# adding portland employment rates from https://data.bls.gov/pdq/SurveyOutputServlet
portland_employment_data <- unemployment_data %>% 
  filter(year >= 2019) %>% 
  filter(cbsa == "76750") %>% 
   group_by(year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_YoY = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100) %>% 
  mutate(avg_emp_YoY = round(avg_emp_YoY, digits = 3))

@@

```{r employment visuals, fig.align = "center", fig.show="hold", fig.height= 7, fig.width=10, eval = FALSE}
# average employment from 2000 to 2022
employment_summary_since2000 <- MSA_employment %>% 
  group_by(MSA) %>% 
  filter(year >= 2000) %>% 
  summarize(
    avg_employment = mean(employment_rate),
    sd_employment = sd(employment_rate)
  )

# visual of MSA employment rates, shows major drop during covid
ggplot(MSA_employment_yoy, aes(x = year, y = employment_rate, color = MSA)) +
  geom_line() +
  xlim(2021, 2022) +
  theme(legend.position = "none") +

@@

```{r payroll data, eval = FALSE}
# payroll data has not been added to final table
payroll_data <- read_csv(here("data","raw", "employee_payroll_MSA21_22.csv")) %>% 
  row_to_names(4) %>% 
  clean_names() %>% 
  rename(MSA = na) %>% 
  rename(payroll_change_may21 = may_2021) %>% 
  rename(payroll_change_mar22 = mar_2022) %>% 
  rename(payroll_change_apr22 = apr_2022) %>% 
  rename(payroll_change_may22 = may_2022_p)

MSA_payroll <- payroll_data %>% 
  full_join(potential_MSAs) %>% 
  drop_na() %>% 
  pivot_longer(cols = 2:5,
               names_prefix = "payroll_change_",
               names_to = "year",
               values_to = "payroll_change") %>% 
  mutate(payroll_change = as.numeric(payroll_change)) %>% 
  group_by(MSA, year)  %>% 
  mutate(payroll_pos = payroll_change >= 0)

```{r payroll visual, fig.height = 7, eval = FALSE}
ggplot(MSA_payroll, aes(x = payroll_change, y = reorder(MSA, payroll_change), fill = payroll_pos)) +
  geom_col(position = "identity", color = "black", size = 0.25) +
  scale_fill_manual(values = c("red", "darkolivegreen3"), guide = FALSE) +
  labs(x = "percent change",
       y = "",
       title = "Quarterly change in employee payroll from 2021 to 2022") +
  facet_grid(~year)

@@

Median housing prices by MSA were obtained from the Association of Realtors for years 2019, 2020, and 2021. The percent difference in 2021 Q1 and 2022 Q1 was used to show the change in price in 2021. **Next quarterly release will be 11/10**

@@

metro_house_prices <- metro_house_cost_data %>%
  select(cbsa, house_price_2021) %>% 
  right_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021, median_income, avg_emp_YoY, median_rent2021, median_rent2022, house_price_2021) %>% 
  filter(popestimate2021 >= 300000)
```

>>>>

### Population Growth
All states were considered with the exception of CA, HI, and AK. Initial analysis of Metropolitan Statistical Areas (MSAs) relied on Census data of annual population estimates, in which the population had to be greater than 300,000 individuals for the purpose of attracting investors. Population change from 2020 to 2021 was calculated by $$\frac{2021 estimate- 2020 estimate}{2020 estimate} * 100$$
MSAs with a population growth > 0.5%, were selected to gather further data on. The 0.5% growth was determined.

@@

```{r population growth data wrangling}
# read in the data MSA population data from Census
msa_pop_2021_data <- read_csv(here("data","raw", "msa_2021_pop_data.csv")) %>% 
  clean_names() %>% # data wrangling to make column headers in lower_snake_case
  filter(lsad == "Metropolitan Statistical Area") %>% # select MSA cities
  select("cbsa", "name", "popestimate2020", "popestimate2021", "residual2020", "residual2021") %>% # select data

  separate(col = "name",
           into = c("MSA", "state"), # split city and state data
           sep = ", ") %>% 
  mutate(pop_percent_change = (popestimate2021-popestimate2020)/popestimate2020 * 100)  # add population % change column

# create object with viable MSAs 
potential_MSAs <- msa_pop_2021_data %>% 
  filter(pop_percent_change > 0.5 ) %>% 
  filter(popestimate2021 >= 300000) %>% 
  mutate(pop_percent_change20_21 = round(pop_percent_change, digits = 3)) %>% 
  filter (!state %in% c("CA", "HI", "AK")) %>% 
  #separate(col = "state", into = c("state1", "state2", "state3"),    sep = "-") %>% 
  filter(!cbsa %in% c("10580", "10900", "12420", "12940", "14260", "15980", "18580", "19100", "19660", "19740", "23540", "25420", "26420", "27260", "29820", "34940", "36260", "36740", "38060", "39100", "39300", "39340", "41620", "45220", "45300", "46060", "46140", "39900")) %>%  # remove MSAs as instructed by Owen 
  mutate(normalized_pop2021 = (popestimate2021-min(popestimate2021))/ (max(popestimate2021)-min(popestimate2021))) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021)

@@

```{r population change visual, fig.height = 7, fig.align='center', eval = FALSE}
ggplot(data = positive_MSA_pop_change, aes(x = pop_percent_change20_21, y = reorder(MSA, pop_percent_change20_21))) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_minimal(base_size = 12.5)

@@

Dataset: Median Household Income (USD) INCOME IN THE PAST 12 MONTHS (IN 2021 INFLATION-ADJUSTED DOLLARS)

@@

```{r household income data wrangling}
MSA_income_data <- read_csv(here("data","raw", "2021_income_data.csv")) %>% 
  row_to_names(1) %>% 
  clean_names() %>% 
  mutate(median_income = as.numeric(estimate_households_median_income_dollars)) %>% 
  mutate(income_moe_USD = as.numeric(margin_of_error_households_median_income_dollars)) %>% 
  separate(col = "geography",
           into = c("id", 'cbsa'),
           sep = 9) %>% 
  mutate(cbsa = as.numeric(cbsa)) # change column class from chr to dbl

@@

avg_emp_yoy <- MSA_employment %>% 
  group_by(MSA, year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_YoY = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100) %>% 
  mutate(avg_emp_YoY = round(avg_emp_YoY, digits = 4)) %>% 
  drop_na()

yoy_test <- MSA_employment %>% 
  group_by(MSA, year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  filter(year >= "2020") %>% 
  pivot_wider(names_from = year,
              values_from = avg_employment_rate,
              names_prefix = "avg_employment_rate") %>% 
  mutate(avg_emp_yoy = (avg_employment_rate2021 - avg_employment_rate2020)/avg_employment_rate2020 *100)

@@

```{r}
# adding portland employment rates from https://data.bls.gov/pdq/SurveyOutputServlet
portland_employment_data <- unemployment_data %>% 
  filter(year >= 2019) %>% 
  filter(cbsa == "76750") %>% 
   group_by(year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_YoY = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100) %>% 
  mutate(avg_emp_YoY = round(avg_emp_YoY, digits = 4))

@@

```{r employment visuals, fig.align = "center", fig.show="hold", fig.height= 7, fig.width=10, eval = FALSE}
# visual of MSA employment rates, shows major drop during covid
ggplot(MSA_employment_yoy, aes(x = year, y = employment_rate, color = MSA)) +
  geom_line() +
  xlim(2021, 2022) +
  theme(legend.position = "none") +

@@

@@

Median housing prices by MSA were obtained from the Association of Realtors for 2021.

@@

metro_house_prices <- metro_house_cost_data %>%
  select(cbsa, house_price_2021) %>% 
  right_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021, median_income, avg_emp_YoY, median_rent2021, median_rent2022, house_price_2021) %>% 
  filter(popestimate2021 >= 500000)
```
